---
title: "Section 5, reproduced: the simulation study"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Section 5, reproduced: the simulation study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 8, fig.height = 5)
has_gg <- requireNamespace("ggplot2", quietly = TRUE)
```

This vignette reproduces the figures and tables of Section 5 of the paper and
its supplement, computed live from the simulation results that ship with the
package. The document is therefore itself a check that the reported numbers
follow from the released runs. The full study can be rerun (about an hour; the
graph-based arms additionally require Python and a Java runtime for py-tetrad)
with

```{r rerun, eval = FALSE}
source(system.file("simulation", "jci_simulation_rebuild.R",
                   package = "directadjust"))
```

## Design

Each replication draws a 15-node DAG with a fixed causal ordering: five latent
variables, seven observed covariates $W_1, \dots, W_7$, then $X_1$, $X_2$, $Y$.
Each node draws its parent count uniformly on
$\{0, \dots, \min(\text{predecessors}, 3)\}$ and its parents uniformly among
predecessors; the treatment edges $X_1 \to Y$ and $X_2 \to Y$ are present in the
effect arm and absent otherwise. Linear-Gaussian data are generated with edge
coefficients uniform on $\pm[0.1, 1.5]$ and error variances uniform on
$[0.5, 1]$, at $n \in \{100; 1{,}000; 10{,}000\}$.

Settings are classified as in the paper: Setting 1 keeps effect-arm models whose
ground-truth tiered PAG (oracle FCI on the true DAG with the four tiers as
background knowledge) contains a GAC-certifiable adjustment set; Setting 2 keeps
effect-arm models where it does not; Setting 3 is the no-effect arm, unscreened.
One hundred models per setting, seed `20260716`.

Four methods run on identical datasets, all at $\alpha = 0.05$, the paper's
testing level: `r1_build()` and `r1_combine()`, and two graph-based arms,
namely FCItiers and GFCI, each supplied with the four tiers as background knowledge and each followed
by the generalized adjustment criterion, with certification computed by dagitty.
Certified sets are validated against the true DAG.

The R1 Combine arm reduces each certified per-treatment set to a minimal
adjustment set through the paper's Lemma 8, the package default
(`prune = "lemma8"`), as Theorem 7 requires. A faster heuristic pruning is
available as `r1_combine(prune = "witness")`; see `?r1_combine`.

```{r load}
library(directadjust)
res <- readRDS(system.file("results", "jci_sim_results.rds",
                           package = "directadjust"))
ARMS <- c("R1 Build", "R1 Combine", "FCItiers + GAC", "GFCI + GAC")
res$method <- as.character(res$method)
res$method[res$method == "FCItiers (Tetrad)"] <- "FCItiers + GAC"
res <- subset(res, method %in% ARMS)
res$method <- factor(res$method, levels = ARMS)
res$outcome <- factor(res$outcome,
  levels = c("correct set", "none correct", "negative", "unknown"))
SETLAB <- c(S1 = "Setting 1", S2 = "Setting 2", S3 = "Setting 3")
```

```{r palette, eval = has_gg}
library(ggplot2)
ORIG_FILL <- scale_fill_manual(values = c(
  "correct set"  = "#0072BD", "none correct" = "#7E7E7E",
  "negative"     = "#D95319", "unknown"      = "#EDB120"),
  labels = c("correct set" = "Correct Set", "none correct" = "None Correct",
             "negative" = "Negative", "unknown" = "Unknown"))
STACK <- geom_bar(position = position_stack(reverse = TRUE))
```

## Figure 6: decision outcomes, Settings 1 and 2

```{r fig6, eval = has_gg}
s12 <- subset(res, setting != "S3")
ggplot(s12, aes(factor(n), fill = outcome)) +
  STACK + ORIG_FILL +
  facet_grid(setting ~ method, labeller = labeller(setting = SETLAB)) +
  theme_minimal() + theme(legend.position = "bottom") +
  labs(x = "sample size", y = "count", fill = NULL)
```

The percentages quoted in Section 5.3 for Setting 1:

```{r sett1-pct}
s1 <- subset(res, setting == "S1")
knitr::kable(
  round(100 * tapply(s1$outcome == "correct set",
                     list(s1$method, s1$n), mean)),
  caption = "Share of the 100 datasets in which a correct set was found (%).")
```

## Figure 7: estimation error, Settings 1 and 2

Whenever a method returns sets, its estimate averages the sum of the two
treatment coefficients from the regression of $Y$ on both treatments and the
set, over all returned sets; negative decisions enter at 0.

```{r fig7, eval = has_gg}
est <- subset(s12, !is.na(est)); est$diff <- est$est - est$true
ggplot(est, aes(factor(n), diff)) +
  geom_boxplot() +
  facet_grid(setting ~ method, labeller = labeller(setting = SETLAB)) +
  theme_minimal() +
  labs(x = "sample size", y = "estimated - true effect")
```

## Mean squared error

One squared error per replication against the model's known true joint effect,
averaged per cell; runs without an estimate are excluded. These are the figures
quoted in the estimation paragraph of Section 5.3.

```{r mse}
est_all <- subset(res, !is.na(est))
est_all$sqerr <- (est_all$est - est_all$true)^2
knitr::kable(
  round(tapply(est_all$sqerr,
               list(paste(SETLAB[as.character(est_all$setting)],
                          format(est_all$n, big.mark = ",", trim = TRUE)),
                    est_all$method), mean), 3),
  caption = "MSE of the joint-effect estimate, by setting and sample size.")
```

## Figure 8: running time against the number of covariates

Median seconds per dataset at $n = 1{,}000$, for the full pipeline (graph
discovery followed by adjustment set enumeration) and for graph discovery alone.
The graphical approaches must enumerate the adjustment sets of the learned graph
before an effect can be estimated; our methods do not.

```{r fig8, eval = has_gg}
sw <- readRDS(system.file("results", "jci_fig8_times.rds",
                          package = "directadjust"))
sw$method <- as.character(sw$method)
sw$method[sw$method == "FCItiers (Tetrad)"] <- "FCItiers + GAC"
sw <- subset(sw, method %in% ARMS)
sw$method <- factor(sw$method, levels = ARMS)
sm <- do.call(rbind, by(sw, list(sw$nW, sw$method), function(d)
  data.frame(nW = d$nW[1], method = d$method[1],
             search = median(d$secs_search), total = median(d$secs_total))))
long <- rbind(
  data.frame(sm[c("nW", "method")], stage = "Full pipeline", secs = sm$total),
  data.frame(sm[c("nW", "method")], stage = "Discovery only", secs = sm$search))
long$stage <- factor(long$stage, levels = c("Full pipeline", "Discovery only"))
ggplot(long, aes(nW, secs, color = method, linetype = stage)) +
  geom_line() + geom_point() +
  scale_linetype_manual(values = c("Full pipeline" = "solid",
                                   "Discovery only" = "22")) +
  scale_x_continuous(breaks = sort(unique(long$nW))) +
  theme_minimal() +
  labs(x = "number of covariates",
       y = "median seconds per dataset (n = 1,000)",
       color = NULL, linetype = NULL)
```

## Supplement: conditional independence test counts

The appendix table. Counts are exact for every arm: the R methods wrap their
test functions with counters, and the Tetrad arms count through a proxy
implementing Tetrad's `IndependenceTest` interface.

```{r ntests}
nt <- subset(res, n == 10000 & !is.na(ntests))
tc <- do.call(rbind, by(nt, list(nt$method, nt$setting), function(d)
  data.frame(Setting = SETLAB[[as.character(d$setting[1])]],
             Method  = as.character(d$method[1]),
             Min = min(d$ntests),
             Q1 = round(quantile(d$ntests, .25)),
             Median = median(d$ntests),
             Q3 = round(quantile(d$ntests, .75)),
             Max = max(d$ntests))))
tc <- tc[order(tc$Setting, match(tc$Method, ARMS)), ]
knitr::kable(tc, row.names = FALSE,
             caption = "Tests per dataset at n = 10,000, over 100 models.")
```

## Supplement: running time distributions

```{r fig-runtime, eval = has_gg}
ggplot(res, aes(factor(n), secs)) +
  geom_boxplot() + scale_y_log10() +
  facet_grid(setting ~ method, labeller = labeller(setting = SETLAB)) +
  theme_minimal() +
  labs(x = "sample size", y = "seconds per dataset (log scale)")
```

## Supplement: Setting 3 (no treatment effect)

```{r fig1011, eval = has_gg}
s3 <- subset(res, setting == "S3")
ggplot(s3, aes(factor(n), fill = outcome)) +
  STACK + ORIG_FILL + facet_grid(. ~ method) +
  theme_minimal() + theme(legend.position = "bottom") +
  labs(x = "sample size", y = "count", fill = NULL)
est3 <- subset(s3, !is.na(est)); est3$diff <- est3$est - est3$true
ggplot(est3, aes(factor(n), diff)) +
  geom_boxplot() + facet_grid(. ~ method) + theme_minimal() +
  labs(x = "sample size", y = "estimated - true effect")
```

## A note on certification

All shipped results certify adjustment sets with dagitty rather than
`pcalg::gac`, which at the time of writing mishandles amenability for joint
treatments (it passes PAGs in which the second treatment has a circle-start
edge into the outcome). A minimal reproducible example ships as
`inst/simulation/gac_bug_repro.R`.
